Rclean: A Tool for Writing Cleaner, More Transparent Code

### 2020-02-04 joss/paper.Rmd
moves prov disc para to methods and adds future directions

The package's main functions are `clean` and `keep`. When provided a
file path to a script and the name of a result (or a set of results),
`clean` analyzes the script's code and extracts the lines of code
required to produce the results. This code can then be passed to the
`keep` function, which can either write the code to disk or copy the
code to the user's clipboard (if no output file path is supplied) and
the user can paste the code into another location (e.g. a script
editor).

@@

@@

We have created ``Rclean`` to provide a simple, easy to use tool for
scientists who would like help refactoring code. Using ``Rclean`` the
code necessary to produce a specified result (e.g., an object stored
in memory or a table or figure written to disk) can be easily and
*reliably* isolated even when tangled with code for other
results. This functionality is enabled by graph analysis of data
provenance collected from the user's script. This is likely to be a
broadly useful tool as statistical programming becomes more common
across the sciences. Tools, such as this, that make the production of
accessible code will be an important aid for improving scientific
reproducibility.

Potential to integrate with a larger context of producing reproducible
projects.

- style = styler, formatr
- naming = lintr, goodpractice
- workflow = drake
- libraries = trackr, rrtools, packrat
- reproducible examples = reprex
- data = pins

Another possible future direction could extend the existing framework
to support new and different provenance capture methods. One
possibility is *retrospective provenance*, which tracks a
computational process as it is executing. Through this active,
concurrent monitoring, retrospective provenance can gather information
that static prospective provenance can't. Only using prospective
provenance means that the outcomes of some processes can not be
predicted. For example, if there is a part of a script that is
determined by a random number, the current implementation of
prospective provenance can not predict the path that will be taken
through the code. Therefore, the code cannot be reduced to exclude the
pathway that would not be taken. However, using retrospective
provenance comes at a cost. In order to gather it, the script needs to
be executed. When scripts are computationally intensive or contain
bugs that stop execution, then retrospective provenance can not be
obtained for part or all of the code. Some work has already been done
in the direction of implementing retrospective provenance for code
cleaning in R (see http://end-to-end-provenance.github.io).

>>>>

The package's main functions are `clean` and `keep`. When provided a
file path to a script and the name of a result (or a set of results),
`clean` analyzes the script's code and extracts the lines of code
required to produce the results. By default, code is formatted
following general best practices recommended by the Tidyverse via the
`styler` package. This code can then be passed to the `keep` function,
which can either write the code to disk or copy the code to the user's
clipboard (if no output file path is supplied) and the user can paste
the code into another location (e.g. a script editor).

@@

In the future, it would also be useful to extend the existing
framework to support other provenance methods. One such possibility is
*retrospective provenance*, which tracks a computational process as it
is executing. Through this active, concurrent monitoring,
retrospective provenance can gather information that static
prospective provenance can't. Greater details of the computational
process would enable other features that could address some
challenges, such as processing information from comments, parsing
control statements and replicating random processes. However, using
retrospective provenance comes at a cost. In order to gather it, the
script needs to be executed. When scripts are computationally
intensive or contain bugs that stop execution, then retrospective
provenance can not be obtained for part or all of the code. Some work
has already been done in the direction of implementing retrospective
provenance for code cleaning in R (see http://end-to-end-provenance.github.io).

@@

We have created ``Rclean`` to provide a simple, easy to use tool for
scientists who would like help refactoring code. Using ``Rclean`` the
code necessary to produce a specified result (e.g., an object stored
in memory or a table or figure written to disk) can be easily and
*reliably* isolated even when tangled with code for other
results. This functionality is enabled by graph analysis of data
provenance collected from the user's script. This is likely to be a
broadly useful tool as statistical programming becomes more common
across the sciences. Tools, such as this, that make it easier to
produce transparent, accessible code will be an important aid for
improving scientific reproducibility.

Although the workflow described is impactful, we see promise in
extending it to interface with other clean code and reproducibility
tools. One example is the `reprex` package, which provides a simple
API for sharing reproducible examples. Another possibility is to help
transtion scripts to function, package and workflow creation and
refactoring via toolboxes like `drake`. `Rclean` could provide a
reliable way to extract parts of a larger script that would be piped
to a simplified reproducible example, in the case of `reprex`, or,
since it can isolate the code from inputs to one or more outputs, be
used to extract all of the components needed to write one or more
functions that would be a part of a package or workflow, as is the
goal of `drake`.
